---
title: "Delayed Plumage Maturation Evolution in Manakins (Aves: Pipridae)"
author: Liam U. Taylor
output: github_document

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = "README.md",
      envir = globalenv()
    )
  })
---

```{r setup, include=FALSE}

setwd("C://Users/Liam/Documents/EEB654/finalproject")

library(tidyverse)
library(tools)
library(ape)
library(geiger)
library(phytools)
library(gridGraphics)
library(cowplot)

STAGES_FILE <- "Data/stages_all.csv"
PLUMAGE_DIR <- "Data/Plumages/"
TREE_FILE <- "Data/pipridae_jetz_10k_consensus.nex"

PLUMAGE_STATE_CODES <- c("Absent"=0,
                         "Carotenoid"=1,
                         "Melanin"=2,
                         "Structural"=3)

```

### Plumage Characters
Wrangling stage characters into a nexus file.
```{r wrangle stages, include=TRUE}

# A function to convert binary characters into probability distributions,
#   so that we can map missing data (NA) to a 50/50 probability across two columns
probMap <- function(v) {
  if (is.na(v)) {
    return(0.5)
  } 
  return(v)
}

# Read in the raw data matrix and concatenate taxa names
stagesData <- read_csv(STAGES_FILE) %>%
          unite(Taxon, Genus, Species, sep="_") %>%
          mutate(S1Binary = as.numeric(Stages>0),
                 S2Binary = as.numeric(Stages>1),
                 S3Binary = as.numeric(Stages>2),
                 Coordinated_Present = map_dbl(Coordinated, probMap),
                 Coordinated_Absent = 1 - Coordinated_Present,
                 Concentrated_Present = map_dbl(Concentrated, probMap),
                 Concentrated_Absent = 1 - Concentrated_Present)
```


Wrangling plumage characters into a nexus file.
```{r wrangle plumages, include=TRUE}

# A function to parse an individual plumage record into a matrix
extractPlumageFile <- function(file) {
  taxon <- strsplit(file, "\\.")[[1]][1]

  plumage <- read_csv(paste(PLUMAGE_DIR, file, sep=""),
                        col_types = cols(.default = col_character()))

  if (nrow(plumage) == 0) {
    plumage <- bind_rows(plumage, tibble(Character="Crown", S1="Absent"))
  }
  
  plumage <- plumage %>%
          pivot_longer(-Character, 
                       names_to="Stage",
                       values_to="Value") %>%
          unite(CharacterStage, Character, Stage, sep="_") %>%
          arrange(CharacterStage) %>%
          pivot_wider(names_from=CharacterStage, 
                      values_from=Value) %>%
          purrr::map(~ PLUMAGE_STATE_CODES[.]) %>%
          as.list() %>%
          as_tibble() %>%
          mutate(Taxon=taxon, .before=1)
  
  return(plumage)
}


# Construct the plumage character matrix
plumagesData <- list.files(PLUMAGE_DIR) %>%
             purrr::map_df(extractPlumageFile)

# All patches not indicated in the original plumage tables are absent
plumagesData[is.na(plumagesData)] <- 0

```

### Backbone Phylogeny
```{r tree pruning, include=TRUE}

# Read the tree file
tree <- read.nexus(TREE_FILE)

# Refactor old taxon names which are common mislabeled across trees
refactorNames <- read_csv("Data/taxon_refactoring_names.csv")

# A function to conditionally refactor a taxon 
refactorTaxon <- function(old) {
  if (old %in% refactorNames$Taxon_Old) {
    new <- refactorNames %>%
        filter(Taxon_Old == old) %>%
        pull(Taxon_New)
    return(new)
  }
  return(old)
}
refactoredTips <- map_chr(tree$tip.label, refactorTaxon)
tree$tip.label <- refactoredTips

# Determine tips which are in the tree but not the data

treeTips <- tree$tip.label

missingTips_stages <- treeTips[!(treeTips %in% stagesData$Taxon)]
missingTips_plumages <- treeTips[!(treeTips %in% plumagesData$Taxon)]

if (length(setdiff(missingTips_stages, missingTips_plumages)) != 0) {
  stop(paste("You do not have the same tip data available for stage and plumage patches\n",
             "Here are the different tips:\n",
             paste(setdiff(treeTips_stages, treeTips_plumages), collapse="  ")))
}

treePruned <- drop.tip(tree, missingTips_stages)
```

### Ancestral State Estimation: Predefinitive Plumage Stages
Mk Model Selection for Stage Evolution
```{r ace - stage - model selection, include=TRUE}


# Organize vector of stage data
stagesVector <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$Stages[1])
names(stagesVector) <- treePruned$tip.label


# First, we test rates models

# MODEL 1: Equal-rates unordered
fit_ER <- fitMk(tree=treePruned, x=stagesVector, model="ER")

# MODEL 2: Equal symmetric ordered
mat_eq_sym_ordered <- matrix(c(0,1,0,0,
                               1,0,1,0,
                               0,1,0,1,
                               0,0,1,0),4,4,byrow=TRUE)
fit_eq_sym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_eq_sym_ordered)

# MODEL 3: Equal asymmetric ordered
mat_eq_asym_ordered <- matrix(c(0,1,0,0,
                                2,0,1,0,
                                0,2,0,1,
                                0,0,2,0),4,4,byrow=TRUE)

fit_eq_asym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_eq_asym_ordered)
  
# MODEL 4: Unequal symmetric ordered
mat_un_sym_ordered <- matrix(c(0,1,0,0,
                               1,0,2,0,
                               0,2,0,3,
                               0,0,3,0),4,4,byrow=TRUE)

fit_un_sym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_un_sym_ordered)


# MODEL 5: Unequal asymmetric ordered
mat_un_asym_ordered <- matrix(c(0,1,0,0,
                                2,0,3,0,
                                0,4,0,5,
                                0,0,6,0),4,4,byrow=TRUE)

fit_un_asym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_un_asym_ordered)

# MODEL 6: All rates different
fit_ARD <- fitMk(tree=treePruned, x=stagesVector, model=)



AICs <- setNames(sapply(list(fit_ER,fit_eq_sym_ordered,fit_eq_asym_ordered,
                             fit_un_sym_ordered,fit_un_asym_ordered,fit_ARD),AIC),
                 c("ER","eq_sym_ordered","eq_asym_ordered",
                   "un_sym_ordered","un_asym_ordered","ARD"))
```

Ancestral State Simulation for Stages
```{r ace - stage - simulation, include=TRUE}

maps <- make.simmap(tree=treePruned,
                    x=stagesVector,
                    model=mat_eq_sym_ordered,
                    nsim=100)

summaries <- summary(maps)

colors <- c("0"="black", 
            "1"="green", 
            "2"="red",
            "3"="blue")

{ 
  plot(summaries, colors=colors)
  trash <- sapply(maps, markChanges, sapply(colors, make.transparent,0.05))
}
```
Correlated evolution
```{r correlations - stage vs. social, include=TRUE}

# First, convert stages to binary characters -- presence of each stage
stage1Binary <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$S1Binary) 
names(stage1Binary) <- treePruned$tip.label

stage2Binary <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$S2Binary) 
names(stage2Binary) <- treePruned$tip.label

stage3Binary <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$S3Binary) 
names(stage3Binary) <- treePruned$tip.label

# Now get a matrix of concentrated lek trait coding
concentrated <- map_df(treePruned$tip.label,
                       ~ filter(stagesData, Taxon==.) %>% select(Concentrated_Present, Concentrated_Absent)) %>%
             as.matrix()
names(concentrated) <- treePruned$tip.label


# And a named matrix of coordinated display trait coding
coordinated <- map_df(treePruned$tip.label,
                      ~ filter(stagesData, Taxon==.) %>% select(Coordinated_Present, Coordinated_Absent)) %>%
            as.matrix()
names(coordinated) <- treePruned$tip.label
  
# A function to extract key results from the fitPagel discrete correlation test
tidyFitPagel <- function(x, y, model) {
  results <- fitPagel(tree=treePruned,
                      x=x,
                      y=y,
                      method="fitMk",
                      model=model)
  ret <- tibble(model=model, p=results$P,
                AIC_ind=results$independent.AIC, AIC_dep=results$dependent.AIC) %>%
      mutate(dAIC = AIC_dep - AIC_ind)
  return(ret)
}

corrs_S1_Concentrated <- map_df(c("ER", "SYM","ARD"), ~ tidyFitPagel(stage1Binary, concentrated, .)) %>%
                      mutate(Social="Concentrated", .before=1) %>%
                      mutate(Stage="S1", .before=2)

corrs_S2_Concentrated <- map_df(c("ER", "SYM","ARD"), ~ tidyFitPagel(stage2Binary, concentrated, .)) %>%
                      mutate(Social="Concentrated", .before=1) %>%
                      mutate(Stage="S2", .before=2)

corrs_S3_Concentrated <- map_df(c("ER", "SYM","ARD"), ~ tidyFitPagel(stage3Binary, concentrated, .)) %>%
                      mutate(Social="Concentrated", .before=1) %>%
                      mutate(Stage="S3", .before=2)

corrs_Concentrated <- bind_rows(corrs_S1_Concentrated, corrs_S2_Concentrated, corrs_S3_Concentrated)


corrs_S1_Coordinated <- map_df(c("ER", "SYM","ARD"), ~ tidyFitPagel(stage1Binary, coordinated, .)) %>%
                     mutate(Social="Coordinated", .before=1) %>%
                     mutate(Stage="S1", .before=2)

corrs_S2_Coordinated <- map_df(c("ER", "SYM","ARD"), ~ tidyFitPagel(stage2Binary, coordinated, .)) %>%
                     mutate(Social="Coordinated", .before=1) %>%
                     mutate(Stage="S2", .before=2)

corrs_S3_Coordinated <- map_df(c("ER", "SYM","ARD"), ~ tidyFitPagel(stage3Binary, coordinated, .)) %>%
                     mutate(Social="Coordinated", .before=1) %>%
                     mutate(Stage="S3", .before=2)

corrs_Coordinated <- bind_rows(corrs_S1_Coordinated, corrs_S2_Coordinated, corrs_S3_Coordinated)

corrs <- bind_rows(corrs_Concentrated, corrs_Coordinated)
```



















### Ancestral State Estimation: Predefinitve Plumage Patches
Produce simmaps
```{r ace - plumage, include=FALSE}

colors <- c("0"="gray", 
            "1"="red", 
            "2"="black",
            "3"="blue")

# A function to generate simmap estimates for a given plumage character
simCharacter <- function(character) {
  chars <- map_dbl(treePruned$tip.label,
                   ~ filter(plumagesData, Taxon==.) %>% pull(!!character))
  
  print(character)
  chars
  names(chars) <- treePruned$tip.label
  
  capture.output( maps <- make.simmap(tree=treePruned,
                       x=chars,
                       model="ARD",
                       nsim=20), 
                  file='NUL')
  return(maps)
}

plumageCharacters <- colnames(plumagesData[,-1])

plumageMaps <- purrr::map(plumageCharacters, simCharacter)

```

Plot simmaps
```{r ace - plumage plots, include=FALSE}

colors <- c("0"="gray", 
            "1"="red", 
            "2"="black",
            "3"="blue")

plotCharacter <- function(maps) {
  
  summaries <- summary(maps)
  plot(summaries, colors=colors, ftype="off")
  
  ret <- recordPlot()
  return(ret)
}

mapPlots <- purrr::map(plumageMaps, plotCharacter)

plot_grid(plotlist = mapPlots, labels=plumageCharacters)

```


## References
Alberch, P., S. J. Gould, G. F. Oster, and D. B. Wake. 1979. Size and shape in ontogeny and phylogeny. Paleobiology:296–317.

Anciães, M., A. Nemésio, and F. Sebaio. 2005. A case of plumage aberration in the Pin-tailed Manakin. Cotinga 23:39–43.

Aramuni, F. V. 2019. Custo individual, ontogenia e visita de fêmeas como moduladores da exibição de corte de machos em *Corapipo gutturalis* (Aves: Pipridae). Thesis for the Instituto Nacional de Pesquisas da Amazônia.

Cárdenas‐Posada, G., C. D. Cadena, J. G. Blake, and B. A. Loiselle. 2018. Display behaviour, social organization and vocal repertoire of Blue-backed Manakin Chiroxiphia pareola napensis in northwest Amazonia. Ibis 160:269–282.

Doucet, S. M., D. B. McDonald, M. S. Foster, and R. P. Clay. 2007. Plumage development and molt in Long-tailed Manakins (*Chiroxiphia linearis*): variation according to sex and age. The Auk 124:29–43.

DuVal, E. H. 2005. Age-based plumage changes in the lance-tailed manakin: a two-year delay in plumage maturation. The Condor 107:915–920.

Foster, M. S. 1987. Delayed maturation, neoteny, and social system differences in two manakins of the genus *Chiroxiphia*. Evolution 41:547–558.

Gaiotti, M. G. 2016. *Antilophia bokermanni* (Aves: Pipridae): Parâmetros reprodutivos, sistema de acasalamento social e genético e o papel da seleção sexual. Thesis for the Universidade de Brasília.

Gould, S. J. 1977. Ontogeny and phylogeny. Harvard University Press.

Hellmayr, C. E. 1929. A contribution to the ornithology of northeastern Brazil. Field Museum of Natural History 12.

Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444–448.

Lewis P.O. 2001. A Likelihood Approach to Estimating Phylogeny from Discrete Morphological Character Data. Systematic Biology.

Mallet-Rodrigues, F., and R. Dutra. 2012. Acquisition of definitive adult plumage in male Blue Manakins *Chiroxiphia caudata*. Cotinga 34.

Marini, M. Â., and R. B. Cavalcanti. 1992. Mating System of the Helmeted Manakin (*Antilophia galeata*) in Central Brazil. The Auk 109:911–913.

Prum, R. O. 1986. The displays of the White-throated Manakin *Corapipo gutturalis* in Suriname. Ibis 128:91–102.

Prum, R. O., and J. Dyck. 2003. A hierarchical model of plumage: morphology, development, and evolution. Journal of Experimental Zoology Part B: Molecular and Developmental Evolution 298:73–90.

Rosselli, L. 1994. The annual cycle of the White-ruffed Manakin *Corapipo leucorrhoa*, a tropical frugivorous altitudinal migrant, and its food plants. Bird Conservation International 4:143–160.
