---
title: "Delayed Plumage Maturation Evolution in Manakins (Aves: Pipridae)"
author: Liam U. Taylor
output: github_document

knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = "README.md",
      envir = globalenv()
    )
  })
---
[Note: This README file is produced by the fully-automated RMarkdown file, `finalproject.rmd` located in the base directory of this repository. Knit that file to reproduce this one.]

# Introduction and Goals
Organisms with deferred reproduction trade the opportunity of current reproduction for the opportunity of development while young [@stearns1992]. In birds with delayed plumage maturation (DPM), this youthful period of reproductive delay is accompanied by a drab "predefinitive" plumage during the periods when older individuals are molting into the colorful "definitive" plumages that characterize increased reproductive effort and success. Research on DPM has largely focused on functional hypotheses for the evolution of predefinitive plumage signals (i.e., three hypotheses on why it is "good to be green"). For example, the "crypsis" hypothesis suggests that drab predefinitive plumages help young males avoid predators [@selander1965; @rohwer1978] while the "social signalling" hypothesis suggests that predefinitive plumages allow young males to reduce the likelihood or costs of predation by signalling reduced reproductive capabilities [@selander1965; @lyon1986]. Studies on these function predefinitive hypotheses are limited on two fronts. First, direct benefits to instantaneous survival rate are insufficient for positive selection on the evolution of deferred reproduction, and are thus insufficient for positive selection on the "delayed maturation" part of "delayed plumage maturation" (Taylor in prep.).

Second, very little research has taken a phylogenetic view of DPM evolution. There are only two such studies. The first phylogenetic paper, by Chu [-@chu1994], found evidence that the evolution predefinitive plumages in various shorebird taxa (Aves: Charadriiformes) was the result of the evolution of juvenile plumage and the retention of an ancestral partial molt. The result is that these shorebirds retain part of their juvenile plumage in their first breeding season. This study suggests that the evolution of DPM was the result of selection on and constraints in molt timing, as opposed to selection on predefinitive plumage patches. The second phylogenetic paper, by Hill [-@hill1996], found that the presence of DPM in *Haemorhous* finches is associated with a reduction in plumage patch size. This study recommends that reduced plumage patches in these finches are themselves associated with limits in carotenoid availability, and thus that the evolution of DPM is associated with an increase in the resource costs of plumage coloration. Unlike the (conceptually limited) functional hypotheses for the evolution of predefinitive plumage signals, phylogenetic investigations have supported constraint-based hypotheses for the evolution of predefinitive plumages associated with developmental shifts.

Manakins (Aves:Passeriformes:Pipridae) are neotropical lekking birds known for their extravagant male plumages and gymnastic sexual displays [@kirwan2011]. Manakin species also vary widely in both the presence and duration of DPM ..... ... .. .. .. . .

```{r plumage examples, include=TRUE}
library(kableExtra)
predefExample <- read_csv("Data/Plumages/Chiroxiphia_linearis.csv")
kbl(predefExample) %>%
  kable_classic(full_width=FALSE, font_size=20, html_font="Arial")
```

# Methods and Results

```{r setup, include=FALSE}

setwd("C://Users/Liam/Documents/EEB654/finalproject")

library(tidyverse)
library(tools)
library(ape)
library(geiger)
library(phytools)
library(gridGraphics)
library(cowplot)

STAGES_FILE <- "Data/stages_all.csv"
PLUMAGE_DIR <- "Data/Plumages/"
TREE_FILE <- "Data/pipridae_jetz_10k_consensus.nex"

PLUMAGE_STATE_CODES <- c("Absent"=0,
                         "Carotenoid"=1,
                         "Melanin"=2,
                         "Structural"=3)

```

### Plumage Characters
Wrangling stage characters into a nexus file.
```{r wrangle stages, include=TRUE}

# A function to convert binary characters into probability distributions,
#   so that we can map missing data (NA) to a 50/50 probability across two columns
probMap <- function(v) {
  if (is.na(v)) {
    return(0.5)
  } 
  return(v)
}

# Read in the raw data matrix, concatenate taxa names,
#   convert multistate stages to binaries for correlation tests,
#   and convert unknown social states to uniform priors across two columns
stagesData <- read_csv(STAGES_FILE) %>%
          unite(Taxon, Genus, Species, sep="_") %>%
          mutate(S1Binary = as.numeric(Stages>0),
                 S2Binary = as.numeric(Stages>1),
                 S3Binary = as.numeric(Stages>2),
                 Coordinated_Present = map_dbl(Coordinated, probMap),
                 Coordinated_Absent = 1 - Coordinated_Present,
                 Concentrated_Present = map_dbl(Concentrated, probMap),
                 Concentrated_Absent = 1 - Concentrated_Present)
```


Wrangling plumage characters into a nexus file.
```{r wrangle plumages, include=TRUE}

# A function to parse an individual plumage record into a matrix
extractPlumageFile <- function(file) {
  taxon <- strsplit(file, "\\.")[[1]][1]

  plumage <- read_csv(paste(PLUMAGE_DIR, file, sep=""),
                        col_types = cols(.default = col_character()))

  if (nrow(plumage) == 0) {
    plumage <- bind_rows(plumage, tibble(Character="Crown", S1="Absent"))
  }
  
  plumage <- plumage %>%
          pivot_longer(-Character, 
                       names_to="Stage",
                       values_to="Value") %>%
          unite(CharacterStage, Character, Stage, sep="_") %>%
          arrange(CharacterStage) %>%
          pivot_wider(names_from=CharacterStage, 
                      values_from=Value) %>%
          purrr::map(~ PLUMAGE_STATE_CODES[.]) %>%
          as.list() %>%
          as_tibble() %>%
          mutate(Taxon=taxon, .before=1)
  
  return(plumage)
}


# Construct the plumage character matrix
plumagesData <- list.files(PLUMAGE_DIR) %>%
             purrr::map_df(extractPlumageFile)

# All patches not indicated in the original plumage tables are absent
plumagesData[is.na(plumagesData)] <- 0

```


### Backbone Phylogeny
```{r tree pruning, include=TRUE}

# Read the tree file
tree <- read.nexus(TREE_FILE)

# Refactor old taxon names which are common mislabeled across trees
refactorNames <- read_csv("Data/taxon_refactoring_names.csv")

# A function to conditionally refactor a taxon 
refactorTaxon <- function(old) {
  if (old %in% refactorNames$Taxon_Old) {
    new <- refactorNames %>%
        filter(Taxon_Old == old) %>%
        pull(Taxon_New)
    return(new)
  }
  return(old)
}
refactoredTips <- map_chr(tree$tip.label, refactorTaxon)
tree$tip.label <- refactoredTips

# Determine tips which are in the tree but not the data

treeTips <- tree$tip.label

missingTips_stages <- treeTips[!(treeTips %in% stagesData$Taxon)]
missingTips_plumages <- treeTips[!(treeTips %in% plumagesData$Taxon)]

if (length(setdiff(missingTips_stages, missingTips_plumages)) != 0) {
  stop(paste("You do not have the same tip data available for stage and plumage patches\n",
             "Here are the different tips:\n",
             paste(setdiff(treeTips_stages, treeTips_plumages), collapse="  ")))
}

treePruned <- drop.tip(tree, missingTips_stages)
```

```{r tree plot, include=FALSE}

colors <- c("0"="black", 
            "1"="green", 
            "2"="red",
            "3"="blue")

# Organize vector of stage data
stagesVector <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$Stages[1])
names(stagesVector) <- treePruned$tip.label


png(filename="Figures/raw_tree.png", width=6, height=8, unit="in", res=400)
dotTree(treePruned, factor(stagesVector), colors=colors, fsize=1.4, ftype="i", length=10)
dev.off()

```


### Ancestral State Estimation: Predefinitive Plumage Stages
Mk Model Selection for Stage Evolution
```{r ace - stage - model selection, include=TRUE}

# First, we test rates models

# MODEL 1: Equal-rates unordered
fit_ER <- fitMk(tree=treePruned, x=stagesVector, model="ER")

# MODEL 2: Equal symmetric ordered
mat_eq_sym_ordered <- matrix(c(0,1,0,0,
                               1,0,1,0,
                               0,1,0,1,
                               0,0,1,0),4,4,byrow=TRUE)
fit_eq_sym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_eq_sym_ordered)

# MODEL 3: Equal asymmetric ordered
mat_eq_asym_ordered <- matrix(c(0,1,0,0,
                                2,0,1,0,
                                0,2,0,1,
                                0,0,2,0),4,4,byrow=TRUE)

fit_eq_asym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_eq_asym_ordered)
  
# MODEL 4: Unequal symmetric ordered
mat_un_sym_ordered <- matrix(c(0,1,0,0,
                               1,0,2,0,
                               0,2,0,3,
                               0,0,3,0),4,4,byrow=TRUE)

fit_un_sym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_un_sym_ordered)


# MODEL 5: Unequal asymmetric ordered
mat_un_asym_ordered <- matrix(c(0,1,0,0,
                                2,0,3,0,
                                0,4,0,5,
                                0,0,6,0),4,4,byrow=TRUE)

fit_un_asym_ordered <- fitMk(tree=treePruned, x=stagesVector, model=mat_un_asym_ordered)


AICs <- setNames(sapply(list(fit_ER,fit_eq_sym_ordered,fit_eq_asym_ordered,
                             fit_un_sym_ordered,fit_un_asym_ordered),AIC),
                 c("ER","eq_sym_ordered","eq_asym_ordered",
                   "un_sym_ordered","un_asym_ordered"))
```

Ancestral State Simulation for Stages
```{r ace - stage - simulation, include=TRUE}
colors <- c("0"="black", 
            "1"="green", 
            "2"="red",
            "3"="blue")

maps <- make.simmap(tree=treePruned,
                    x=stagesVector,
                    model=mat_eq_sym_ordered,
                    nsim=1000)
summaries <- summary(maps)

png(filename="Figures/ace_tree.png", width=6, height=8, unit="in", res=100)
plot(summaries, colors=colors)
trash <- sapply(maps, markChanges, sapply(colors, make.transparent,0.01))
dev.off()
```
### Ancestral State Estimation: Predefinitive Plumage Patches

```{r ace - plumage - ace, include=FALSE}
# A function to generate simmap estimates for a given plumage character
simCharacter <- function(character) {
  chars <- map_dbl(treePruned$tip.label,
                   ~ filter(plumagesData, Taxon==.) %>% pull(!!character))
  
  # Convert to binary
  chars[chars > 0] <- 1
  names(chars) <- treePruned$tip.label
  
  if (all(chars==0)) {
    return(NA)
  }
  maps <- make.simmap(tree=treePruned,
                      x=chars,
                      model="ARD",
                      nsim=100)
  return(maps)
}

mapsPlumage <- purrr::map(plumageCharacters, simCharacter)
names(mapsPlumage) <- plumageCharacters
mapsPlumage <- mapsPlumage[!is.na(mapsPlumage)]
```

Plot simmaps
```{r ace - plumage plots, include=TRUE}
colors <- c("0"="gray", 
            "1"="red")

png(filename="Figures/ace_plumages.png", width=15, height=15, unit="in", res=50)
par(mfrow=c(6,6))
for (i in 1:length(mapsPlumage)) {
  densityMap(mapsPlumage[[i]], colors=colors, ftype="off", mar=c(3,3,3,0), legend=FALSE)
  title(main=plumageCharacters[i], cex.main=3)
}
dev.off()
```

```{r ace - plumage - transitions, include=FALSE}

png(filename="Figures/sims_crown_s1.png", width=4, height=4, unit="in", res=400)
densityMap(mapsPlumage[["Crown_S1"]], colors=colors, ftype="off", mar=c(3,3,3,0), legend=FALSE)
title(main="Crown_S1", cex.main=2)
dev.off()

png(filename="Figures/sims_face_s2.png", width=4, height=4, unit="in", res=400)
densityMap(mapsPlumage[["Face_S2"]], colors=colors, ftype="off", mar=c(3,3,3,0), legend=FALSE)
title(main="Face_S2", cex.main=2)
dev.off()

```

Correlated evolution

```{r correlations - stage vs. social, include=TRUE}

# First, convert stages to binary characters -- presence of each stage
stage1Binary <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$S1Binary) 
names(stage1Binary) <- treePruned$tip.label

stage2Binary <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$S2Binary) 
names(stage2Binary) <- treePruned$tip.label

stage3Binary <- map_dbl(treePruned$tip.label,
                        ~ filter(stagesData, Taxon==.)$S3Binary) 
names(stage3Binary) <- treePruned$tip.label

# Now get a matrix of concentrated lek trait coding
concentrated <- map_df(treePruned$tip.label,
                       ~ filter(stagesData, Taxon==.) %>% select(Concentrated_Present, Concentrated_Absent)) %>%
             as.matrix()
rownames(concentrated) <- treePruned$tip.label

# And a named matrix of coordinated display trait coding
coordinated <- map_df(treePruned$tip.label,
                      ~ filter(stagesData, Taxon==.) %>% select(Coordinated_Present, Coordinated_Absent)) %>%
            as.matrix()
rownames(coordinated) <- treePruned$tip.label

```

```{r correlations - stage vs. social, include=TRUE}
colors <- c("0"="white",
            "1"="gray",
            "2"="black")

multiTreeBinary <- map_df(treePruned$tip.label,
                           ~ filter(stagesData, Taxon==.) %>% select(S2Binary, Coordinated, Concentrated)) %>%
                as.matrix()
rownames(multiTreeBinary) <- treePruned$tip.label

multiTreeBinary <- multiTreeBinary + 1
multiTreeBinary[is.na(multiTreeBinary)] <- 0

png(filename="Figures/social_tree.png", width=6, height=6, unit="in", res=100)
dotTree(tree=treePruned, x=multiTreeBinary, colors=colors, data.type="discrete", ftype="reg", legend=FALSE)
dev.off()
```



```{r correlations - stage vs. social - Pagel test, include=TRUE}

# Perform an Mk likelihood comparison for ER vs ARD models with the binary characters
#   NOTE we cannot fit the model for the social characters themselves, because of missing data
#     (even though the fitPagel method can account for that missing data as a flat prior)

# A function to extract the AIC score from a fitMk run
tidyFitMk <- function(chars, model) {
  fit <- fitMk(tree=treePruned, x=chars, model=model)
  ll <- fit$logLik
  AIC <- AIC(fit)
  
  ret <- tibble(Model=model, lhood=ll, AIC=AIC)
  return(ret)
}

fitStage1Binary <- map_df(c("ER", "ARD"), ~ tidyFitMk(stage1Binary, .)) %>%
                mutate(Character="Stage1Binary", .before=1)

fitStage2Binary <- map_df(c("ER", "ARD"), ~ tidyFitMk(stage2Binary, .)) %>%
                mutate(Character="Stage2Binary", .before=1)

fitStage3Binary <- map_df(c("ER", "ARD"), ~ tidyFitMk(stage3Binary, .)) %>%
                mutate(Character="Stage3Binary", .before=1)

fitsStagesBinary <- bind_rows(fitStage1Binary, fitStage2Binary, fitStage3Binary) %>%
                 group_by(Character) %>%
                 filter(AIC == min(AIC))
```


```{r correlations - stage vs. social - Pagel test, include=TRUE}

dependencyKey <- function(d) {
  if (d == "x") {
    return("X<-Y")
  } else if (d == "y") {
    return("X->Y")
  } else if (d == "xy") {
    return ("X<->Y")
  }
  return("ERROR")
}

# A function to extract key results from the fitPagel discrete correlation test
tidyFitPagel <- function(x, y, model, dep) {
  results <- fitPagel(tree=treePruned,
                      x=x,
                      y=y,
                      method="fitMk",
                      dep.var=dep,
                      model=model)
  ret <- tibble(model=model, p=results$P,
                AIC_ind=results$independent.AIC, AIC_dep=results$dependent.AIC) %>%
      mutate(dAIC = AIC_dep - AIC_ind,
             Dependency = map_chr(dep, dependencyKey))
  return(ret)
}

# Gone fishing
corrs_S1_Concentrated <- map_df(c("xy", "x", "y"), ~ tidyFitPagel(stage1Binary, concentrated, "ER", .)) %>%
                      mutate(Social="Concentrated", .before=1) %>%
                      mutate(Stage="S1", .before=2)

corrs_S2_Concentrated <- map_df(c("xy", "x", "y"), ~ tidyFitPagel(stage2Binary, concentrated, "ER", .)) %>%
                      mutate(Social="Concentrated", .before=1) %>%
                      mutate(Stage="S2", .before=2)

corrs_S3_Concentrated <- map_df(c("xy", "x", "y"), ~ tidyFitPagel(stage3Binary, concentrated, "ER", .)) %>%
                      mutate(Social="Concentrated", .before=1) %>%
                      mutate(Stage="S3", .before=2)

corrs_Concentrated <- bind_rows(corrs_S1_Concentrated, corrs_S2_Concentrated, corrs_S3_Concentrated)


corrs_S1_Coordinated <- map_df(c("xy", "x", "y"), ~ tidyFitPagel(stage1Binary, coordinated, "ER", .)) %>%
                      mutate(Social="Coordinated", .before=1) %>%
                      mutate(Stage="S1", .before=2)

corrs_S2_Coordinated <- map_df(c("xy", "x", "y"), ~ tidyFitPagel(stage2Binary, coordinated, "ER", .)) %>%
                      mutate(Social="Coordinated", .before=1) %>%
                      mutate(Stage="S2", .before=2)

corrs_S3_Coordinated <- map_df(c("xy", "x", "y"), ~ tidyFitPagel(stage3Binary, coordinated, "ER", .)) %>%
                      mutate(Social="Coordinated", .before=1) %>%
                      mutate(Stage="S3", .before=2)

corrs_Coordinated <- bind_rows(corrs_S1_Coordinated, corrs_S2_Coordinated, corrs_S3_Coordinated)

corrs <- bind_rows(corrs_Concentrated, corrs_Coordinated)
```

```{r correlations - tables, echo=TRUE}
coevolutions <- corrs %>%
             filter(Dependency == "X<->Y") %>%
             select(-model)

kbl(coevolutions, digits=2) %>%
  kable_classic(full_width=FALSE, font_size=20, html_font="Arial")

sigCorrs <- corrs %>%
         filter(Dependency != "X<->Y", p <= 0.05) %>%
         select(-model)

kbl(sigCorrs, digits=2) %>%
  kable_classic(full_width=FALSE, font_size=20, html_font="Arial")
```


# Discussion and Conclusion

# References



## References
Alberch, P., S. J. Gould, G. F. Oster, and D. B. Wake. 1979. Size and shape in ontogeny and phylogeny. Paleobiology:296–317.

Anciães, M., A. Nemésio, and F. Sebaio. 2005. A case of plumage aberration in the Pin-tailed Manakin. Cotinga 23:39–43.

Aramuni, F. V. 2019. Custo individual, ontogenia e visita de fêmeas como moduladores da exibição de corte de machos em *Corapipo gutturalis* (Aves: Pipridae). Thesis for the Instituto Nacional de Pesquisas da Amazônia.

Cárdenas‐Posada, G., C. D. Cadena, J. G. Blake, and B. A. Loiselle. 2018. Display behaviour, social organization and vocal repertoire of Blue-backed Manakin Chiroxiphia pareola napensis in northwest Amazonia. Ibis 160:269–282.

Doucet, S. M., D. B. McDonald, M. S. Foster, and R. P. Clay. 2007. Plumage development and molt in Long-tailed Manakins (*Chiroxiphia linearis*): variation according to sex and age. The Auk 124:29–43.

DuVal, E. H. 2005. Age-based plumage changes in the lance-tailed manakin: a two-year delay in plumage maturation. The Condor 107:915–920.

Foster, M. S. 1987. Delayed maturation, neoteny, and social system differences in two manakins of the genus *Chiroxiphia*. Evolution 41:547–558.

Gaiotti, M. G. 2016. *Antilophia bokermanni* (Aves: Pipridae): Parâmetros reprodutivos, sistema de acasalamento social e genético e o papel da seleção sexual. Thesis for the Universidade de Brasília.

Gould, S. J. 1977. Ontogeny and phylogeny. Harvard University Press.

Hellmayr, C. E. 1929. A contribution to the ornithology of northeastern Brazil. Field Museum of Natural History 12.

Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444–448.

Lewis P.O. 2001. A Likelihood Approach to Estimating Phylogeny from Discrete Morphological Character Data. Systematic Biology.

Mallet-Rodrigues, F., and R. Dutra. 2012. Acquisition of definitive adult plumage in male Blue Manakins *Chiroxiphia caudata*. Cotinga 34.

Marini, M. Â., and R. B. Cavalcanti. 1992. Mating System of the Helmeted Manakin (*Antilophia galeata*) in Central Brazil. The Auk 109:911–913.

Prum, R. O. 1986. The displays of the White-throated Manakin *Corapipo gutturalis* in Suriname. Ibis 128:91–102.

Prum, R. O., and J. Dyck. 2003. A hierarchical model of plumage: morphology, development, and evolution. Journal of Experimental Zoology Part B: Molecular and Developmental Evolution 298:73–90.

Rosselli, L. 1994. The annual cycle of the White-ruffed Manakin *Corapipo leucorrhoa*, a tropical frugivorous altitudinal migrant, and its food plants. Bird Conservation International 4:143–160.
